How to publish in Biological Sciences: a guide for the uninitiated

### 2021-12-15 4.03-problempeerreview.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

There is already a lot covering peer review in this book, and I have placed this chapter last not because it is the least significant, potentially it is the most significant, but because I think that it is important that you appreciate exactly what peer review is, and to some extent experience it, before you begin to consider the problems with the peer review system.

@@

If peer review was effective, then reviews of grant applications should closely align with the productivity of grants given. Fang et al. [-@fang2016nih] found that percentile scores awarded by peer review of NIH grant applications were poor at predicting the the productivity of >100 000 grants awarded.

@@

One of the shocking results of a very large study of peer review of _PLOS ONE_ articles is the large number of comments that are written directly attacking the authors as a group or personally [i.e. _ad hominem_ attacks, see @eve2021reading]. This should not happen. Reviewers should be confining their objective comments to the work and its presentation. However, this is an aspect of peer review where authors (especially the corresponding and leading authors) will need to acquire a thick skin, because unprofessional comments are made to people across gender and racial groupings, but especially toward traditionally underrepresented groups [@silbiger2019unprofessional]. Sadly, these same groups feel that such comments disproportionately impact their productivity and career advancement [@silbiger2019unprofessional]. Reading comments that are sent to other authors can be cathartic as these allow you to see that everyone receives such negative comments. [ShitMyReviewers](https://shitmyreviewerssay.tumblr.com/) is a good source of these, or see Eve et al. [-@eve2021reading], or Silbiger and Stubler [-@silbiger2019unprofessional]. When _ad hominem_ attacks are made, it would be good if editors openly and explicitly identified these as **bad behaviour**. It would certainly improve the understanding of authors if editors intervened when such _ad hominem_ attacks are made. This would not necessarily involve deleting these comments, but directing authors to ignore the same.

@@

| Bias for which there is evidence | Study demonstrating bias | 
| - | - | 
| Against female authors | @tregenza2002gender; @manlove2018authors; @fox2019gender; @budden2008double-blind; @morgan2018foundation; @hagan2020women |
| Against female reviewers | @helmer2017gender; @fox2019gendera |
| Towards author reputation, favouring acceptance of manuscripts despite poor reviews | @bravo2018hidden; @okike2016single-blind |
| Towards authors from more prestigious institutions, also called prestige bias | @ceci1982peer; @travis1991new; @tomkins2017single; @manlove2018authors ; @lee2013bias |
| Nationality and language bias | @lee2013bias; @manlove2018authors; @nunez2021monolingual; @link1998us |
| Confirmation bias (the tendency for journals and reviewers to favour significant results) | @mahoney1977publication; @fanelli2010positive; @fanelli2012negative; [see Part 1](transparency.html) |
| Publication bias (the literature contains a bias in published results) | @munafo2007association; @dongen2011associations; @fanelli2017meta-assessment; @franco2014publication; [see Part 4](#badscience) |

@@

The truth is that all reviews are not equal because some reviewers will put in more effort than others. Some will know the literature better. Some will be experts in the field that should be better placed to comment. These people are actually more likely to be less senior, PhD students, post docs or early career researchers. However, the importance for the editor is not to take account of the names of these people, their rank, their institution, or other demographics such as their gender, race or nationality. There are great editors out there who can do this, but my impression is that the majority fail. In this case, the only way to do this is for the [triple blind method](#triple). Here the editors will invite the reviewers (by name) but the reviews that result will not be marked with the reviewers’ names. This will make forgetting who they are easier, especially for busy editors.

@@

Today, editors are so busy with the other duties in their jobs as academics that their decisions are hurried and expecting them to take the time and space to overcome their personal biases might be a lot to ask. Instead, I think that it is time for the [triple blind review](#triple) concept to move into the mainstream so that editors can more easily not be led by potential biases of their reviewers, and instead be swayed by the quality of the review and their own reading of the manuscript.

@@

This evidence that peer review is a collaborative system towards the betterment of science is, to me, a sign that all is alive and well in science, and that peer review is acting as it should. However with any social network comes the fragilities of human bias. This means that while peer review may function well for some, for others it may more often than not fail. The bigger problem is that it might depend on your sex, the colour of your skin, the name of your institution, or your country as to whether you are selected as a potential reviewer (i.e. to join the club; see Table \@ref(tab:Bias)), or having submitted your manuscript, find that peer review is going to work for you [see @davies2021promoting]. In addition, if you are never asked to review then you will never benefit from this network.

@@

There are lots of ways in which to do improve editorial oversight. My intuition is that the crux is to find a way that makes it more efficient and objective for the editors. For example, to try to pin down reviewers on where they find fault and exactly what that fault is. There is a difference between:

@@

Hence, fixing peer review comes back to fixing problems associated with the publishing culture (and all that that entails), rather than any a simple fix-all for the myriad of existing publishing options. [Preprints](#preprints) combined with [Overlay journals](#overlay) offer one solution that keeps reviews with the original submission. This stops that practice of authors resubmitting their manuscripts over and over to countless journals until a pair of reviewers fail to spot a fundamental flaw. Keeping these manuscripts with their reviews on BiorXiv represents one solution, together with another problem that good articles with biased reviews still need to overcome. I feel that this is more likely to come right when authors resubmit to another Overlay journal with a valid rebuttal.

Lastly, but perhaps most importantly, we have to be more realistic about the limitations of peer review. We would be better to think of peer review as a '**silver standard**' - something that some scientists agree has merit. Our problems come because our expectations of peer review are too great - a 'gold standard' it is not. Instead, we can keep the 'gold standard' for those papers that have withstood the test of time, the repeatability of the community, and acceptance into the mainstream.

>>>>

There is already a lot covering peer review in this book, and I have placed this chapter last not because it is the least significant, potentially it is the most significant, but because I think that it is important that you appreciate exactly what peer review is, and experience it from both sides, before you begin to consider the problems with the peer review system.

@@

For example, if peer review was effective, then reviews of grant applications should closely align with the productivity of grants given. Fang et al. [-@fang2016nih] found that percentile scores awarded by peer review of NIH grant applications were poor at predicting the the productivity of >100 000 grants awarded.

@@

One of the shocking results of a very large study of peer review of _PLOS ONE_ articles is the large number of comments that are written directly attacking the authors as a group or personally [i.e. _ad hominem_ attacks, see @eve2021reading]. This should not happen. Reviewers should confine their objective comments to the work and its presentation. However, this is an aspect of peer review where authors (especially the corresponding and leading authors) will need to acquire a thick skin, because unprofessional comments are made to people across gender and racial groupings, but especially toward traditionally underrepresented groups [@silbiger2019unprofessional]. Sadly, these same groups feel that such comments disproportionately impact their productivity and career advancement [@silbiger2019unprofessional]. Reading comments that are sent to other authors can be cathartic as these allow you to see that everyone receives such negative comments. [ShitMyReviewers](https://shitmyreviewerssay.tumblr.com/) is a good source of these, or see Eve et al. [-@eve2021reading], or Silbiger and Stubler [-@silbiger2019unprofessional]. When _ad hominem_ attacks are made, it would be good if editors openly and explicitly identified these as **bad behaviour**. It would certainly improve the understanding of authors if editors intervened when such _ad hominem_ attacks are made. This would not necessarily involve deleting these comments, but directing authors to ignore the same.

@@

| Bias for which there is evidence | Study demonstrating bias | 
| - | - | 
| Against female authors | @tregenza2002gender; @manlove2018authors; @fox2019gender; @budden2008double-blind; @morgan2018foundation; @hagan2020women |
| Against female reviewers | @helmer2017gender; @fox2019gendera |
| Towards author reputation, favouring acceptance of manuscripts despite poor reviews | @bravo2018hidden; @okike2016single-blind |
| Towards authors from more prestigious institutions, also called prestige bias | @ceci1982peer; @travis1991new; @garfunkel1994effect; @tomkins2017single; @manlove2018authors ; @lee2013bias |
| Nationality and language bias | @song2000publication; @lee2013bias; @manlove2018authors; @nunez2021monolingual; @link1998us |
| Confirmation bias (the tendency for journals and reviewers to favour significant results) | @mahoney1977publication; @fanelli2010positive; @fanelli2012negative; [see Part I](transparency.html) |
| Publication bias (the literature contains a bias in published results) | @jennions2002publication; @munafo2007association; @dongen2011associations; @franco2014publication; @fanelli2017meta-assessment; @sanchez-tojar2018meta-analysis; [see Part IV](#badscience) |

@@

The truth is that all reviews are not equal because some reviewers will put in more effort than others. Some will know the literature better. Some will be experts in the field that should be better placed to comment. These people are actually more likely to be less senior, PhD students, post docs or early career researchers. However, the importance for the editor is not to take account of the names of these people, their rank, their institution, or other demographics such as their gender, race or nationality. There are great editors out there who can do this, but my impression is that the majority fail. In this case, the only way to do this is by the [triple blind method](#triple). Here the editors will invite the reviewers (by name) but the reviews that result will not be marked with the reviewers’ names. This will make forgetting who they are easier, especially for busy editors.

@@

Today, editors are so busy with the other duties in their jobs as academics that their decisions are hurried and expecting them to take the time and space to overcome their personal biases might be a lot to ask. Instead, I think that it is time for the [Open Evaluation](#publicreviewers) concept to move into the mainstream so that everyone can see how editors came by their decision and were not led by potential biases of their reviewers, and instead be swayed by the quality of the review and their own reading of the manuscript. This is especially important for rejected manuscripts, which is why we need the effort of this peer review recorded on [preprint](#preprints) sites - such as happens in [overlay journals](#overlay).

@@

This evidence that peer review is a collaborative system towards the betterment of science is, to me, a sign  that peer review is acting as it should. However with any social network comes the fragilities of human bias. This means that while peer review may function well for some, for others it may more often than not fail. The bigger problem is that it might depend on your sex, the colour of your skin, the name of your institution, or your country as to whether you are selected as a potential reviewer (i.e. to join the club; see Table \@ref(tab:Bias)), or having submitted your manuscript, find that peer review is going to work for you [see @davies2021promoting]. In addition, if you are never asked to review then you will never benefit from this network.

@@

There are lots of ways by which editorial oversight can be improved. My intuition is that the crux is to find a way that makes it more efficient and objective for the editors. For example, to try to pin down reviewers on where they find fault and exactly what that fault is. There is a difference between:

@@

Hence, fixing peer review comes back to fixing problems associated with the publishing culture (and all that that entails), rather than any a simple fix-all for the myriad of existing publishing options. [Preprints](#preprints) combined with [Overlay journals](#overlay) offer one solution that keeps reviews with the original submission. This stops that practice of authors resubmitting their manuscripts over and over to countless journals until a pair of reviewers fail to spot a fundamental flaw. Keeping these manuscripts with their reviews on *biorXiv* (or any other preprint server) represents one solution, together with another problem that good articles with biased reviews still need to overcome. I feel that this is more likely to come right when authors resubmit to another Overlay journal with a valid rebuttal.

### Peer review is here to stay

Lastly, but perhaps most importantly, we have to be more realistic about the limitations of peer review. We would be better to think of peer review as a '**silver standard**' - something that some scientists agree has merit. Our problems come because our expectations of peer review are too great - a 'gold standard' it is not. An invaluable filter that improves the quality of manuscripts through a spirit of professional camaraderie - it is (for the most part). We can keep the 'gold standard' for those papers that have withstood the test of time, the repeatability of the community, and acceptance into the mainstream.

### 2021-12-15 4.04-predatory.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

Predatory journals are publications that purport to be from scholarly publishing houses, but have little or no editorial oversight or peer review. They exist in order to extract the Article Processing Charge (APC) that is so ubiquitous in [Open Access](#openaccess) journals. They continue to exist because publishing is moving towards APCs, and there is little difference between what they do and some supposedly ‘legitimate’ journals. For example, some definitions of predatory journals include that their APCs far exceed their publishing costs, but this can now certainly be said of some legitimate journals.

## If the line is so grey, how is it possible to tell whether or not a journal is predatory?
This is a surprisingly difficult question to answer. Predatory journals have become so sophisticated at what they do, that it can be very difficult to determine whether or not they are legitimate. Moreover, their electronically published journals can create new titles faster than the time that it takes to check that they are legitimate. An example of a publisher in the grey zone is Hindawi, that were once considered predatory but were later removed from Beall’s list. At the time of writing, I considered that Hindawi still had some questionable practices and so suggested continuing to avoid them. However, in a new (January 2021) twist, Wiley (a well respected publisher) [acquired Hindawi for USD 298 million @michael2021wiley] (yes folks, your science is big business because of the big profits that Wiley expect to make out of you and your tax paying funders). This news goes to show how quickly the publishing world is reacting to the new world of Open Access (limitless numbers of titles), but also how some publishers intend to dominate the market and maintain their stranglehold on scientific publishing.

Most academics have an interaction with predatory journals through email. Indeed, the numbers of emails that academics are spammed with is astounding [@vandenberg2021solution], and while the best advice is to use a spam filter, I still find that too many legitimate emails go there too. These spamming emails from predatory publishers are not straight forward and seek to manipulate the reader into submitting a manuscript [@bett2020predatory]. Moreover, predatory journals target the most vulnerable in the community, poorer researchers who do not have English as their first language are deliberately targeted [@bett2020predatory, @lund2020analysis].

At the same time, legitimate journals have become increasingly predatory in their habits, and it’s difficult to tell them apart from predatory journals. There was a time when it was possible to categorically state that no journal will ever approach you with a general email that invites you to contribute. However, there are now several legitimate journals that send out unsolicited emails. Realistically then, in the current publishing world, there is a continuum from predatory to legitimate. It was not always this way, and that means that as an early career researcher you are facing difficulties not faced by your mentor or other more senior academics. Not only do you need to avoid publishing in predatory journals, but you should also avoid citing their articles.

@@

@@

1. **Use an index**. Web of Science and Scopus both curate contents of legitimate journals. If your journal of choice appears in one of these, then it is very likely to be legitimate. Note that Google Scholar includes many predatory journals, so please never use this to determine whether or not a journal is legitimate. Note also that it takes a journal several years to gain enough kudos to get accepted onto Web of Science and/or Scopus. Therefore, it can still be legitimate and not be there. How to choose the right journal for your publication is covered in an earlier section ([Part 1](#whichjournal)). If the journal you want to publish in is not in Web of Science or Scopus, then proceed to the next step.
1. **Ask your librarian**. Librarians are fantastic sources as well as custodians of information, and journals are one of their key knowledge areas. Don’t hesitate to get in touch with your librarian and ask their advice. They are likely to be very well placed to respond to your request. They may also be guardians of granting APCs at your institution, so it is in their interest to make sure that these valuable monies don’t fall into the wrong hands.
1. **Ask your mentor or an experienced colleague**. It’s worth doing this with them so that you can see the steps that they follow. Given that steps 1 and 2 have already come back with uncertain answers, spreading your net more widely will help with step 4. However, be warned that there are increasing numbers of senior scientists that have been caught out by predatory journals, so checking their contributors is now no cut and dry way to differentiate between them and legitimate journals.
1. **Who is on the editorial board?** Journals publish names of their editors, associate editors and the editorial board. Look through these lists and see whether there are names that you recognise. If you know any of the people, you (or your advisor) can contact and ask them about the journal (they should be happy to respond). Be warned that it is easy to place someone’s name on a website, so unless they have personally told you, keep away.
1. **Check against a known list**. In the past, this might have been the first thing to do, but the number of predatory journals is proliferating so quickly that it’s hard for any list to keep up. Beall’s list retired in 2017. The next best list now has more than three times the number of journal names. A new list, by Simon Linacre, is sadly behind a paywall, so you can’t expect to access it. One of the reasons why Beall gave up is that the new tactic for these publishers is to produce lots of new journals. Curating a list is real work and has implications for the publishers on it, hence you now need to pay to access an up to date list. There are more lists: Cabell's Predatory Journal Blacklist, Beall’s list, DOAJ delisted journals, Scopus discontinued sources, etc.

@@

The first step would be to write to the publisher and withdraw the article. Whether or not you paid an APC, having it on their website is not good for your reputation. Beware, these journals don’t adhere to an ethical code, and so they might refuse to withdraw your paper. Or they may want to charge an additional fee to remove it (remember that they are in it for the money).

>>>>

Predatory journals are publications that purport to be from scholarly publishing houses, but have little or no editorial oversight or peer review. They exist in order to extract the Article Processing Charge ([APC](#APC)) that is so ubiquitous in [Open Access](#openaccess) journals. They continue to exist because for-profit publishing is moving towards [APC](#APC)s, and there is little difference between what they do and some supposedly ‘legitimate’ journals. For example, some definitions of predatory journals include that their [APC](#APC)s far exceed their publishing costs, but this can now certainly be said of many legitimate journals [@grossmann2021current].

> “Predatory journals and publishers are entities that prioritize self-interest at the expense of scholarship and are characterized by false or misleading information, deviation from best editorial and publication practices, a lack of transparency, and/or the use of aggressive and indiscriminate solicitation practices.”
>
> ---@grudniewicz2019predatory

## If the line is so grey, how is it possible to tell whether or not a journal is predatory?
This is a surprisingly difficult question to answer. Predatory journals have become so sophisticated at what they do, that it can be very difficult to determine whether or not they are legitimate. Some have managed to get themselves listed on the curated databases like PubMed and Scopus [@manca2020predatory, @cortegiani2020citations]. Moreover, their electronically published journals can create new titles faster than the time that it takes to check that they are legitimate. An example of a publisher in the grey zone is Hindawi, that were once considered predatory but were later removed from Beall’s list. At the time of writing, I considered that Hindawi still had some questionable practices and so suggested continuing to avoid them. However, in a new (January 2021) twist, Wiley (a respected publisher) [acquired Hindawi for USD 298 million @michael2021wiley] (yes folks, your science is big business because of the big profits that Wiley expect to make out of you and your tax paying funders). This news goes to show how quickly the publishing world is reacting to the new world of Open Access (limitless numbers of titles), but also how some publishers intend to dominate the market and maintain their stranglehold on scientific publishing.

Most academics have an interaction with predatory journals through email. Indeed, the numbers of emails that academics are spammed with is astounding [@vandenberg2021solution], and while the best advice is to use a spam filter, I still find that too many legitimate emails go there too. These spamming emails from predatory publishers are not straight forward and seek to manipulate the reader into submitting a manuscript [@bett2020predatory]. Moreover, predatory journals target the most vulnerable in the community, poorer researchers who do not have English as their first language are deliberately targeted [@bett2020predatory; @lund2020analysis].

At the same time, legitimate journals have become increasingly predatory in their habits, and it’s difficult to tell them apart from predatory journals [@crosetto2021mdpi; @oviedo-garcia2021journal]. There was a time when it was possible to categorically state that no journal will ever approach you with a general email that invites you to contribute. However, there are now several legitimate journals that send out unsolicited emails. Realistically then, in the current publishing world, there is a continuum from predatory to legitimate.

A pertinent example is the Multidisciplinary Digital Publishing Institute (MDPI: originally included on Beall’s list and later removed). For a thorough investigation of MDPI see Ángeles Oviedo-García [-@oviedo-garcia2021journal], and Critchlow and Stefenelli [-@critchlow2021comment] for the publisher's response. MDPI is pertinent because they are currently an ascending star in [Gold OA](#gold) publishing endeavours in terms of their volume, and in addition they are one of the most aggressive in their marketing strategies [@crosetto2021mdpi; @oviedo-garcia2021journal]. MDPI has a mixture of titles with diverse quality that rub along together. However, the publishers are committed to income from [Gold OA](#gold) with a [no Impact Factor model](#noIF). This led them into direct conflict with the [editorial board](#boards) of one title, _Nutrients_, who did not want to allow low quality articles to be published [@devrieze2018open-access]. Editors started to receive comments from MDPI staff complaining that too many papers were being rejected. Is this the action of a predatory journal worried about income being lost? Or a publisher who advocates a [no Impact Factor model](#noIF)? Like other aspects of modern scholarly publishing, the overriding thirst for income from publishers makes it very hard to draw the line between legitimate and predatory.

For the MDPI title _Nutrients_, a mass resignation ensued, and this is the really one of the only ways that editorial boards have of showing their displeasure with a publisher; see also the story of [_Glossa_](#gatekeepers) and _Diversity and Distributions_ [@peterson2019open]. However, the standing problem is always that there appear to be plenty more academics to take their places. MDPI might be in more trouble as at least one university in the Czech Republic has told their staff that they will not support any further payments to MDPI, and will not regard anything published there as a scholarly output. Time will tell whether other institutes join them. But if we consider the practices of MDPI to be predatory, shouldn't we also extend this to other large for-profit publishers?

It was not always this way, and that means that as an Early Career Researcher you are facing difficulties not faced by your [mentor](#mentor) or other more senior academics. Not only do you need to avoid publishing in predatory journals, but you should also avoid citing their articles.

@@

### Sometimes the predators are caught {#OMICS}
In 2019, a US judge fined OMICS International Publishing $50 million for the deception of authors that paid for publication of submitted manuscripts in their predatory journals, or attended bogus conferences [@brainard2019US]. She found that less than half of the manuscripts had been sent out for peer review. OMICS also refused to retract any articles after authors found out that they had been duped. The $50 million represented total company revenue for the years 2011 to 2017.

@@

1. **Use an index**. Web of Science and Scopus both curate contents of legitimate journals. If your journal of choice appears in one of these, then it is very likely to be legitimate. Note that Google Scholar includes many predatory journals, so please never use this to determine whether or not a journal is legitimate. Note also that it takes a journal several years to gain enough kudos to get accepted onto Web of Science and/or Scopus. Therefore, it can still be legitimate and not be there. How to choose the right journal for your publication is covered in an earlier section ([Part I](#whichjournal)). If the journal you want to publish in is not in Web of Science or Scopus, then proceed to the next step.
1. **Ask your librarian**. Librarians are fantastic sources as well as custodians of information, and journals are one of their key knowledge areas. Don’t hesitate to get in touch with your librarian and ask their advice. They are likely to be very well placed to respond to your request. They may also be guardians of granting [APC](#APC)s at your institution, so it is in their interest to make sure that these valuable monies don’t fall into the wrong hands.
1. **Ask your mentor or an experienced colleague**. It’s worth doing this with them so that you can see the steps that they follow. Given that steps 1 and 2 have already come back with uncertain answers, spreading your net more widely will help with step 4. However, be warned that there are increasing numbers of senior scientists that have been caught out by predatory journals, so checking their contributors is now no cut and dry way to differentiate between them and legitimate journals.
1. **Who is on the editorial board?** Journals publish names of their editors, associate editors and the editorial board. Look through these lists and see whether there are names that you recognise. If you know any of the people, you (or your advisor) can contact and ask them about the journal (they should be happy to respond). Be warned that it is easy to place someone’s name on a website, so unless they have personally told you, keep away.
1. **Check against a known list**. In the past, this might have been the first thing to do, but the number of predatory journals is proliferating so quickly that it’s hard for any list to keep up. Beall’s list retired in 2017 (although a version with updates is maintained here). The next best list now has more than three times the number of journal names. A new list, by Simon Linacre, is sadly behind a paywall, so you can’t expect to access it. One of the reasons why Beall gave up is that the new tactic for these publishers is to produce lots of new journals. Curating a list is real work and has implications for the publishers on it, hence you now need to pay to access an up to date list. There are more lists: Cabell's Predatory Journal Blacklist, Beall’s list, DOAJ delisted journals, Scopus discontinued sources, [ThinkCheckSubmit](https://thinkchecksubmit.org/), etc.

@@

The first step would be to write to the publisher and withdraw the article. Whether or not you paid an [APC](#APC), having it on their website is not good for your reputation. Beware, these journals don’t adhere to an ethical code, and so they might refuse to withdraw your paper. Or they may want to charge an additional fee to remove it (remember that they are in it for the money).
